

# Introduction

# Data description

The data represents a subset of samples from a larger experiment and consists of
in total 10 mzML files with 3 samples from individuals with a cardiovascular
disease, 3 samples from healthy donors and 4 QC samples (pool of all samples of
the experiment). The MS data has been restricted to a retention time range from
20 to 230 seconds and an m/z range from 0 to 500.



# Data pre-processing

The *mzML* files with the raw MS data are located within the *data/mzML* folder
of this repository.

```{r}
library(xcms)
library(RColorBrewer)
library(pander)

#' read the sample descriptions from a text file
pd <- read.table("data/phenodata.txt", header = TRUE, sep = "\t")

#' Import the data
data <- readMSData(paste0("data/mzML/", pd$mzML_file),
                   pdata = new("AnnotatedDataFrame", pd),
                   mode = "onDisk")
```

```{r phenodata, results = "asis"}
pandoc.table(pData(data), style = "rmarkdown",
             caption = "Samples from the data set.")
```

```{r define-colors}
col_phenotype <- brewer.pal(4, "Set1")[c(2, 1, 4)]
names(col_phenotype) <- c("CTR", "CVD", "QC")
```

## General quality assessment

```{r bpc}
bpc <- chromatogram(data, aggregationFun = "max")
```

```{r}
plot(bpc, col = paste0(col_phenotype[bpc$phenotype], 80))
grid()
legend("topright", col = col_phenotype,
       legend = names(col_phenotype), lty = 1)
```

- BPC and heatmaps to evaluate LC performance.
- EIC of internal standards.

## Chromatographic peak detection

## Retention time alignment

## Correspondence

## Gap filling

# Data normalization

- Evaluation on IS.

# Identification of interesting features

- Pre-filtering.
- Differential abundance.

Task: sensitivity analysis: with/without normalization. with/without
pre-filtering.


# Annotation

- Different levels of annotation.
- Just m/z (mass).
- m/z and retention time.
- MS/MS spectra + public repository.
- MS/MS spectra + retention time.

# Summary

# Session information

# References

