# Testing code and evaluating data

This file contains some code to test how to best analyze the data from the
end-to-end workflow. There's not much of description and also no diagnostic
plots, variance evaluations and visualizations.

## Pre-processing

```{r}
library(xcms)
library(MsExperiment)
library(Spectra)
library(RColorBrewer)
register(MulticoreParam(6))

pd <- data.frame(
    mzML = c("20180622_QC4_POS.mzML", "20180622_ARVD 15_POS.mzML",
             "20180622_CTR21_POS.mzML", "20180622_QC7_POS.mzML",
             "20180622_CTR25_POS.mzML", "20180622_ARVD 8_POS.mzML",
             "20180622_QC8_POS.mzML", "20180622_CTR22_POS.mzML",
             "20180622_ARVD 5_POS.mzML", "20180622_QC9_POS.mzML"),
    group = c("QC", "CVD", "CTR", "QC", "CTR", "CVD", "QC", "CTR", "CVD", "QC"),
    inj_idx = c(45, 50, 51, 78, 82, 85, 89, 93, 94, 99))

mse <- readMsExperiment(paste0("data/mzML_orig/", pd$mzML), sampleData = pd)

col_phenotype <- brewer.pal(4, "Set1")[c(2, 1, 4)]
names(col_phenotype) <- c("CTR", "CVD", "QC")
col_sample <- col_phenotype[sampleData(mse)$group]

bpc <- chromatogram(mse, aggregationFun = "max")
plot(bpc, col = paste0(col_sample, 80), lwd = 2)

mse <- filterRt(mse, c(10, 240))
bpc <- chromatogram(mse, aggregationFun = "max")
plot(bpc, col = paste0(col_sample, 80), lwd = 2)

cwp <- CentWaveParam(peakwidth = c(2, 10), ppm = 40, integrate = 2)
xmse <- findChromPeaks(mse, param = cwp, chunkSize = 5)

rcp <- MergeNeighboringPeaksParam(expandRt = 2.5, expandMz = 0.001,
                                  minProp = 3/4)
xmse <- refineChromPeaks(xmse, param = rcp, chunkSize = 5)

pdp <- PeakDensityParam(sampleGroups = sampleData(xmse)$group,
                        minFraction = 0.75,
                        binSize = 0.01,
                        bw = 2)
xmse <- groupChromPeaks(xmse, param = pdp)

pgp <- PeakGroupsParam(minFraction = 0.9, extraPeaks = 50, span = 0.3,
                       subsetAdjust = "average",
                       subset = which(sampleData(xmse)$group == "QC"))
xmse <- adjustRtime(xmse, param = pgp)
plotAdjustedRtime(xmse)

pdp <- PeakDensityParam(sampleGroups = sampleData(xmse)$group,
                        minFraction = 1/3, binSize = 0.015, bw = 2.3)
xmse <- groupChromPeaks(xmse, param = pdp)

xmse <- fillChromPeaks(xmse, param = ChromPeakAreaParam(), chunkSize = 5)

library(SummarizedExperiment)
se <- quantify(xmse, method = "sum", filled = TRUE)

## impute
set.seed(123)
assays(se)$imputed <- imputeRowMinRand(assay(se), method = "from_to",
                                       min_fraction_from = 1/2,
                                       min_fraction = 1)

```

## Data visualization

For now only a PCA is created - might need also other plots.

```{r}
library(CompMetaboTools)
pc <- prcomp(t(log2(assay(se, "imputed"))),
             scale = FALSE, center = TRUE)
par(mfrow = c(1, 2))
plot_pca(pc, bg = paste0(col_sample, 80), pc_x = 1, pc_y = 2,
         main = "raw data", pch = 21, col = "#00000080")
legend("topleft", col = col_phenotype, legend = names(col_phenotype),
       pch = 16, cex = 0.4)
plot_pca(pc, bg = paste0(col_sample, 80), pc_x = 3, pc_y = 4,
         main = "raw data", pch = 21, col = "#00000080")
```

## Data normalization

Maybe run median scaling here.

## Differential abundance analysis

Use *limma* for differential abundance analysis.

```{r}
library(limma)

se_qc <- se[, se$group == "QC"]
se <- se[, se$group != "QC"]

dratios <- apply(
    log2(assay(se_qc, "imputed")), 1, sd, na.rm = TRUE) /
    apply(log2(assay(se, "imputed")), 1, sd, na.rm = TRUE)

plot(density(dratios, na.rm = TRUE), xlab = "D-ratio",
     main = "Distribution of D-ratios")
abline(v = 0.5, col = "red", lty = 2)

se <- se[which(dratios < 0.5), ]
keep <- moreAreValidThan(assay(se, "raw"), f = se$group, prop = 0.5)
se <- se[keep, ]

pheno <- factor(se$group, levels = c("CTR", "CVD"))
dsgn <- model.matrix(~ 0 + pheno)
fit <- limma::lmFit(log2(assay(se, "imputed")), design = dsgn)

contr_mat <- limma::makeContrasts(
    CVDvsCTR = phenoCVD - phenoCTR,
    levels = dsgn
)
fit <- limma::contrasts.fit(fit, contrasts = contr_mat)
fit <- limma::eBayes(fit)

rowData(se)$coef <- fit$coefficient[, 1]
rowData(se)$pvalue <- fit$p.value[, 1]
rowData(se)$adjp <- p.adjust(fit$p.value[, 1], method = "BH")

save(se, file = "data/se.RData")
plot(rowData(se)$coef, -log10(rowData(se)$adjp))
abline(h = -log10(c(0.05, 0.1, 0.35)))

res <- rowData(se)[rowData(se)$adjp < 0.35, ]
res <- res[order(res$pvalue), ]
res$feature_id <- rownames(res)

res[, c("mzmed", "rtmed", "adjp", "coef")] |> as.data.frame()

```

## Compound annotation

### Based on m/z values of features

Match m/z values against theoretical m/z values for compounds' adducts.

```{r}
library(MetaboAnnotation)
library(AnnotationHub)
ah <- AnnotationHub()

mb <- ah[["AH107049"]]

mz_mtch <- matchValues(
    query = res, target = compounds(mb), mzColname = "mzmed",
    param = Mass2MzParam(adducts = c("[M+H]+", "[M+Na]+")))
```

Add results to the table.

```{r}
res$mz_names <- unlist(lapply(mz_mtch, function(z)
    paste(unique(z$target_name), collapse = "; ")))
res$mz_adduct <- unlist(lapply(mz_mtch, function(z)
    paste(unique(z$adduct), collapse = "; ")))
```

mz 138.0546: Trigonelline (found in coffee)
mz 195.0877: Caffeine

Check which features could be isotopes: need also the intensities of the
features though. Median intensity?

```{r}
#' Get the max intensities for each feature (chrom peak) for the
#' sample with the highest overall signal.
eics <- featureChromatograms(xmse, features = rownames(res))
#' identify the sample with the highest max signal
max_ints <- lapply(seq_len(ncol(eics)), function(i) {
    vapply(eics[, i], function(z) max(intensity(z), na.rm = TRUE), numeric(1))
})
max_ints <- do.call(cbind, max_ints)
max_ints[!is.finite(max_ints)] <- 0
res$intmax <- max_ints[, which.max(colSums(max_ints))]

#' Alternative: calculate the median (integrated!) signal across samples
#' for each feature
res$intmed <- rowMedians(assay(se)[res$feature_id, ], na.rm = TRUE)

plot(log2(res$intmed), log2(res$intmax))
grid()
abline(0, 1)
#' so, they are correlated.

par(mfrow = c(1, 2))
plot(res$mzmed, res$intmed, type = "h")
plot(res$mzmed, res$intmax, type = "h")

## TODO: order x if needed.
library(MetaboCoreUtils)
m <- res[, c("mzmed", "intmed")] |> as.matrix()
m <- m[order(m[, "mzmed"]), ]
isotopologues(m, ppm = 50)
isotopologues(res[, c("mzmed", "intmax")], ppm = 0, tolerance = 0.01)
#' nothing for either one.

```

```{r}
ms1 <- featureSpectra(xmse, msLevel = 1L, features = rownames(res)[1])
```

Isotope search did not find anything.

### Based on m/z values and retention times from previously measured standards

Match using our internal standards library.

```{r}
#' load the data.frame with retention times.

```

### Using fragment (MS/MS) spectra

Load MS/MS data for the present data set.

```{r}
## include the LC-MS/MS data...
fls <- c("20210129_ARVD_QC_CE20_HighSensitivity_POS.mzML",
         "20210129_ARVD_QC_CE30_HighSensitivity_POS.mzML",
         "20210129_ARVD_QC_CE40_HighSensitivity_POS.mzML",
         "20210129_ARVD_QC_MS_HighSensitivity_POS_1.mzML"
         )
ms2 <- filterMsLevel(Spectra(paste0("data/mzML_LCMSMS/", fls)), 2L)

#' Remove peaks above precursor, including precursor.
ms2 <- filterPrecursorPeaks(ms2, ppm = 0, tolerance = 0.05, mz = ">=")
ms2 <- filterIntensity(ms2, intensity = function(z)
    z > max(z, na.rm = TRUE) * 0.01)
ms2 <- ms2[lengths(ms2) >= 2]

int_sum <- function(x, ...) {
    x[, "intensity"] <- x[, "intensity"] / sum(x[, "intensity"], na.rm = TRUE)
    x
}
ms2 <- addProcessing(ms2, int_sum)

#' Find MS2 spectra matching the m/z and retention times of our features
ms2match <- matchValues(res, ms2, param = MzRtParam(ppm = 10, toleranceRt = 5),
                        mzColname = c("mzmed", "precursorMz"),
                        rtColname = c("rtmed", "rtime"))

res_ms2 <- target(ms2match)[targetIndex(ms2match)]
res_ms2$feature_id <- query(ms2match)$feature_id[queryIndex(ms2match)]

```

From which data files are these MS/MS spectra?

```{r}
dataStorage(target(ms2match))[whichTarget(ms2match)] |>
    basename() |>
    table()
```

Match these MS2 spectra with fragment spectra in MassBank.

```{r}
mbs <- Spectra(mb)
mbs <- filterIntensity(mbs, intensity = function(z)
    z > max(z, na.rm = TRUE) * 0.01)
mbs <- mbs[lengths(mbs) >= 2]
mbs <- addProcessing(mbs, int_sum)

#' Identify MassBank spectra with a similarity >= 0.7
register(SerialParam())
ms2mb <- matchSpectra(res_ms2, mbs, param = CompareSpectraParam(ppm = 50))

tmp <- ms2mb[whichQuery(ms2mb)]
tmp <- pruneTarget(tmp)

#' From which mzML file are these?
dataOrigin(query(tmp)) |>
    basename() |>
    table()

#' It's all caffeine!
tmp$target_name
plotSpectraMirror(tmp[1])
plotSpectraMirror(tmp[2])
plotSpectraMirror(tmp[3])
plotSpectraMirror(tmp[4])
plotSpectraMirror(tmp[5])
plotSpectraMirror(tmp[6])

tmp$rtime
```

What with the other MS2 spectra? what could they be? Maybe an ion different than
[M+H]+?

Retry without precursor m/z.

```{r}
res_ms2 <- setBackend(res_ms2, MsBackendMemory())
res_ms2 <- applyProcessing(res_ms2)

mbs <- setBackend(mbs, MsBackendMemory())
mbs <- applyProcessing(mbs)

ms2mb2 <- matchSpectra(
    res_ms2, mbs, param = CompareSpectraParam(
                      ppm = 50, requirePrecursor = FALSE))
ms2mb2
```
